How do changes in ex vary between countries?

### 2022-07-19 draft_working_paper.Rmd
save formatted proposed final images

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

@@

### Changes in life expectancy at age 65

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

@@

>>>>

And a simplified version to save

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country)  +
  labs(
    x = "Year",
    y = "Life Expectancy at birth (years)",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database.\nSynthetic Germany is an 80%/20% weighted average of West Germany and East Germany respectively",
    colour = "Sex", linetype = "Sex"
  ) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave("figures/proposed/e0_trend.png", height = 20, width = 20, units = "cm", dpi = 300)

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

I'm going to make this the first proposed figure

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>%
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65 (years)",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database.\nSynthetic Germany is an 80%/20% weighted average of West Germany and East Germany respectively",
    colour = "Sex", linetype = "Sex"
  ) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave("figures/proposed/e65_trend.png", height = 20, width = 20, units = "cm", dpi = 300)
```

@@

I now want a simplified and slightly better looking version of this for the main manuscript

```{r}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  mutate(delta_ex = delta_ex * 52.25) %>% # Convert to weeks
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  stat_smooth(se = FALSE) + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (in weeks)",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

ggsave("figures/proposed/che0_smoother.png", height = 15, width = 25, units = "cm", dpi = 300)

And now the same but at age 65

```{r}
change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  mutate(delta_ex = delta_ex * 52.25) %>% # Convert to weeks
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  stat_smooth(se = FALSE) + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (in weeks)",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

ggsave("figures/proposed/che65_smoother.png", height = 15, width = 25, units = "cm", dpi = 300)

@@

I'm going to just get the predicted values for the best model

```{r}
best_breakpoint_model_predictions <- 
  segmented_breakpoints_models %>% 
    mutate(bic = map_dbl(mdl_outputs, BIC)) %>% 
    group_by(code, x, sex) %>% 
    filter(bic == min(bic)) %>% 
    mutate(predictions = map2(mdl_outputs, data, predict)) %>% 
    mutate(data_augmented = map2(data, predictions, ~ .x %>% mutate(prediction = .y))) %>% 
    select(code, x, sex, data_augmented) %>% 
    ungroup() %>% 
    unnest(data_augmented)

First I want to show the values against predicted

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = prediction), colour = "red") +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at birth, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

ggsave("figures/proposed/e0_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)

```
Now, at age 65

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = prediction), colour = "red") +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at age 65, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

ggsave("figures/proposed/e65_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)

And now, the same, but differences

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 0) %>% 
  group_by(country, sex) %>% 
  arrange(year) %>% 
  mutate(
    ch_ex      = 52.25 * (ex - lag(ex)                 ), 
    ch_ex_pred = 52.25 * (prediction - lag(prediction) )
  ) %>% 
  filter(!is.na(ch_ex), !is.na(ch_ex_pred))  %>% 
  ggplot(aes(year)) + 
  geom_point(aes(y = ch_ex), alpha = 0.35) +
  facet_grid(sex ~ country) +
  geom_hline(yintercept = 0) + 
  geom_line(aes(y = ch_ex_pred), colour = "red") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (weeks)",
    title = "Predicted and observed annual changes in life expectancy at birth, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

ggsave("figures/proposed/ch_e0_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)

```
And finally for age 65

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 65) %>% 
  group_by(country, sex) %>% 
  arrange(year) %>% 
  mutate(
    ch_ex      = 52.25 * (ex - lag(ex)                 ), 
    ch_ex_pred = 52.25 * (prediction - lag(prediction) )
  ) %>% 
  filter(!is.na(ch_ex), !is.na(ch_ex_pred))  %>% 
  ggplot(aes(year)) + 
  geom_point(aes(y = ch_ex), alpha = 0.35) +
  facet_grid(sex ~ country) +
  geom_hline(yintercept = 0) + 
  geom_line(aes(y = ch_ex_pred), colour = "red") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (weeks)",
    title = "Predicted and observed annual changes in life expectancy at age 65, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

ggsave("figures/proposed/ch_e65_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
